working title Compatibility system and stigma size are the main predictors of heterospecific pollen effect

### 2018-12-19 Manuscript_draft/Manuscript_draft.Rmd
changes

---
title: working title: Compatibility system and stigma size are the main predictors of heterospecific pollen effect
author:
- affilnum: '1,2,3'
  name: Jose B. Lanuza, Ignasi Bartomeus, Tia-Lynn Ashman, Romina Rader
  email: barragansljose@gmail.com
affiliation:
- affil: US Environmental Protection Agency, Office of Research and Development, National
    Health and Environmental Effects Research Laboratory, Atlantic Ecology Division,
    27 Tarzwell Drive  Narragansett, RI, 02882, USA
  affilnum: 1
- affil: Big Name University, Department of R, City, BN, 01020, USA
  affilnum: 2
- affil: Estacion Biologica de Donana (EBD-CSIC), E-41092 Sevilla, Spain
  affilnum: 3
output:
  pdf_document:
    fig_caption: yes
    fig_height: 7
    fig_width: 7
    includes:
      in_header: header.tex
    keep_tex: yes
    number_sections: no
    template: manuscript.latex
  word_document: null
  html_document: null
capsize: normalsize
classoption: a4paper
csl: ecology.csl
documentclass: article
fontsize: 11pt
footerdate: yes
geometry: margin=1in
linenumbers: yes
bibliography:
- references.bib
- knitcitations.bib
spacing: doublespacing
abstract: no
---

@@

library("knitr")

@@

In most ecosystems, plant species normally coexist and share their floral visitors with other species @bascompte2003 [better ref is Waser et al 1996 on generalization: https://esajournals.onlinelibrary.wiley.com/doi/abs/10.2307/2265575](). From the plants’ perspective, pollinator sharing can be positive if there is facilitation amog plant species @carvalheiro2014 or negative due to competition at the pre-pollination stage Pauw (2013) [easier to say will be positive for some plants and negative for others, depending on the facilitation competition gradient](). An increasing number of visits often correlates with higher chances of fertilization @engel2003. However this is not always the case, among these possible flower visitors there are also nectar robbers and pollen thieves @inouye1980 [See also Magrach et al 2017 Nat Ecol Evol]. Receiving both sufficient quantity and quality deposited on the stigma is thus highly relevant to the pollination success of the plant @aizen2007.

@@

By visiting many plant species, many pollinators are responsible for conspecific pollen loss and the transport of foreign pollen, both of which can have important detrimental effects on species fitness @morales2008;  @ashman2013; @arceo2016. Foreign pollen arrival can play an important role in plant species fitness but outcomes are variable and appear to be context dependent as there is not always a decrease in fitness @morales2008. Some of this variation is likely due to the enormous variability of foreign pollen transferred across systems ranging from 0 to 75 percent. However, most studies report ranges of heterospecific pollen between 0 and 20 percent of the total pollen load @bartomeus2008 @montgomery2012; @ashman2013; @fang2013, yet even these relatively low amounts of heterospecific pollen transferred can decrease fitness greatly @thomson1982. While we now have some understanding of the impacts of heterospecific pollen quantity, we have less understanding of other factors that could be driving the variation in impacts upon fitness.  @ashman2013 postulated the first predictive framework that identifies a need to understand how plant traits might mediate heterospecific pollen effect, whereby mating system and pollen size were predicted to potentially mediate the impact of foreign pollen transfer on plant fitness. This concept is supported by specific case studies, such as @tong2016 that demonstrate an asymmetrical effect in 6 species of *Pedicularis* whereby the pollen of long styled species was able to grow the full length of the style on short styled species but not vice versa. While this suggests that the impacts of heterospecific pollen may differ among pollen donor and recipient, few studies have been conducted to ascertain whether this pattern is in fact a general trend or to identify the extent to which other plant traits are critical to heterospecific pollen impacts. [this pargraph is very good!]()

@@

Incompatibility system is another plant trait that appears to play an important role in foreign pollen effect. Species that are self incompatible have stronger barriers to heterospecific pollen than self-compatible species @ashman2013. However, note the large variability in mating systems across populations @whitehead2018 [...](). The type of incompatibility, (i.e. whether sporophytic or gametophytic) is related to the location of pollen recognition; sporophytic incompatibility relates to signaling at the stigma surface while gametophytic occurs within the style @barrett1988.  This later acting pollen recognition mechanism is associated with greater negative effect than sporophytic recognistion @ashman2013.[if you introduce this, I eexpect you will use it in your analysis, remove?]() [Nonetheless, an effect of foreign pollen is a bit obscured by the variability within species, however species that are strong selfers or strong outcrossers have less variablity in mating systems and predictions of effect could be more realistic (see figure 1 from @whitehead2018)- IB this is a side problem for you, mention just once, and maybe in discusion. Now it has a lot of weigth.](). Moreover, other traits such as number of pollen grains per flower and number of ovules have been traditionally associated with the type of incompatibility system where species with higher pollen ovule ratios are predicted to be xenogamous and species with low pollen ovule ratios autogamous (REF) . Selfer species are known to have a reduction of herkogamy (REF) and less pollen production per ovule (REF) which can be interpretated as a reduction of pollen exported into the community [so, pollen and ovules are important by themselfs, or only because correlated? Unclear what your point is](). Other morphological traits, like stigma size can be determinant for the total pollen quatity that a stigma can receive and therefore related to do that pollen size would also play an important role. Example with pollen here.

@@

Foreign pollen effect was studied through two different treatments, one with 50% conspecific pollen and 50% heterospecific pollen and a second one with 100% foreign pollen (N=10) [this second I don't get, maybe explain it's ultility.](). Therefore, 180 different combinations were performed with N=10 per combination. Seed set was the proxy of effect for all our treatments. Moreover, hand cross pollination [100% cross, right? Clarify](), hand self pollination, apomixis (bagged emasculated flowers) and natural selfing were tested for each species (N=10). All?¿ flowers were emasculated the day prior anthesis and hand pollinated next day with a toothpick. Hand-pollination was conducted with 3-4 gentle touches on the stigma surface. [The mixes of pollen were realized on an eppendorf based on the pollen counts made with Neubaeur chamber (each anther was counted 4 times for 20 different anthers per species)-IB explain better and give a bit more of detail](). In order to confirm that the treatments applied were 50-50 percent pollen, for each focal species the total stigmatic load of pollen was counted from one donor of each family (N=3).

@@

>>>>

---
title: working title Compatibility system and stigma size are the main predictors of heterospecific pollen effect
author:
- affilnum: '1,2,3'
  name: Jose B. Lanuza, Ignasi Bartomeus, Tia-Lynn Ashman, Romina Rader
  email: barragansljose@gmail.com
affiliation:
- affil: US Environmental Protection Agency, Office of Research and Development, National
    Health and Environmental Effects Research Laboratory, Atlantic Ecology Division,
    27 Tarzwell Drive  Narragansett, RI, 02882, USA
  affilnum: 1
- affil: Big Name University, Department of R, City, BN, 01020, USA
  affilnum: 2
- affil: Estacion Biologica de Donana (EBD-CSIC), E-41092 Sevilla, Spain
  affilnum: 3
output:
  pdf_document:
    fig_caption: yes
    fig_height: 7
    fig_width: 7
    includes:
      in_header: header.tex
    keep_tex: yes
    number_sections: no
    template: manuscript.latex
  word_document: null
  html_document: null
capsize: normalsize
classoption: a4paper
csl: ecology.csl
documentclass: article
fontsize: 11pt
footerdate: yes
geometry: margin=1in
linenumbers: yes
bibliography:
- references.bib
- knitcitations.bib
spacing: doublespacing
abstract: no
---

@@

library("knitr")
library("ggplot2")
library("extrafont")

@@

In most ecosystems, plant species normally coexist and share their floral visitors with other species @waser1996. From the plants’ perspective, pollinator sharing can be positive for some plants @carvalheiro2014 or negative for others @pauw2013, depending on the facilitation gradient. An increasing number of visits often correlates with higher chances of fertilization @engel2003. However this is not always the case, among these possible flower visitors there are also nectar robbers and pollen thieves @inouye1980; @magrach2017. Receiving both sufficient quantity and quality deposited on the stigma is thus highly relevant to the pollination success of the plant @aizen2007.

@@

By visiting many plant species, many pollinators are responsible for conspecific pollen loss and the transport of foreign pollen, both of which can have important detrimental effects on species fitness @morales2008;  @ashman2013; @arceo2016. Foreign pollen arrival can play an important role in plant species fitness but outcomes are variable and appear to be context dependent as there is not always a decrease in fitness @morales2008. Some of this variation is likely due to the enormous variability of foreign pollen transferred across systems ranging from 0 to 75 percent. However, most studies report ranges of heterospecific pollen between 0 and 20 percent of the total pollen load @bartomeus2008 @montgomery2012; @ashman2013; @fang2013, yet even these relatively low amounts of heterospecific pollen transferred can decrease fitness greatly @thomson1982. While we now have some understanding of the impacts of heterospecific pollen quantity, we have less understanding of other factors that could be driving the variation in impacts upon fitness.  @ashman2013 postulated the first predictive framework that identifies a need to understand how plant traits might mediate heterospecific pollen effect, whereby mating system and pollen size were predicted to potentially mediate the impact of foreign pollen transfer on plant fitness. This concept is supported by specific case studies, such as @tong2016 that demonstrate an asymmetrical effect in 6 species of *Pedicularis* whereby the pollen of long styled species was able to grow the full length of the style on short styled species but not vice versa. While this suggests that the impacts of heterospecific pollen may differ among pollen donor and recipient, few studies have been conducted to ascertain whether this pattern is in fact a general trend or to identify the extent to which other plant traits are critical to heterospecific pollen impacts.

@@

Incompatibility system is another plant trait that appears to play an important role in foreign pollen effect. Species that are self incompatible have stronger barriers to heterospecific pollen than self-compatible species @ashman2013. However, note the large variability in mating systems across populations @whitehead2018 [...](). The type of incompatibility, (i.e. whether sporophytic or gametophytic) is related to the location of pollen recognition; sporophytic incompatibility relates to signaling at the stigma surface while gametophytic occurs within the style @barrett1988.  This later acting pollen recognition mechanism is associated with greater negative effect than sporophytic recognistion @ashman2013. [if you introduce this,I eexpect you will use it in your analysis, remove]() [Nonetheless, an effect of foreign pollen is a bit obscured by the variability within species, however species that are strong selfers or strong outcrossers have less variablity in mating systems and predictions of effect could be more realistic (see figure 1 from @whitehead2018)- IB this is a side problem for you, mention just once, and maybe in discusion. Now it has a lot of weigth.](). Moreover, other traits such as number of pollen grains per flower and number of ovules have been traditionally associated with the type of incompatibility system where species with higher pollen ovule ratios are predicted to be xenogamous and species with low pollen ovule ratios autogamous (REF) . Selfer species are known to have a reduction of herkogamy (REF) and less pollen production per ovule (REF) which can be interpretated as a reduction of pollen exported into the community [so, pollen and ovules are important by themselfs, or only because correlated? Unclear what your point is](). Other morphological traits, like stigma size can be determinant for the total pollen quatity that a stigma can receive and therefore related to do that pollen size would also play an important role. Example with pollen here.

@@

Foreign pollen effect was studied through two different treatments, one with 50% conspecific pollen and 50% heterospecific pollen and a second one with 100% foreign pollen (N=10) [this second I don't get, maybe explain it's ultility.](). Therefore, 180 different combinations were performed with N=10 per combination. Seed set was the proxy of effect for all our treatments. Moreover, hand cross pollination (between individuals of the same species), hand self pollination, apomixis (bagged emasculated flowers) and natural selfing were tested for each species (N=10). For the treatments with foreign pollen and hand cross pollination, flowers were emasculated the day prior anthesis and hand pollinated next day with a toothpick. Hand-pollination was conducted with 3-4 gentle touches on the stigma surface. [The mixes of pollen were realized on an eppendorf based on the pollen counts made with Neubaeur chamber (each anther was counted 4 times for 20 different anthers per species)-IB explain better and give a bit more of detail](). In order to confirm that the treatments applied were 50-50 percent pollen, for each focal species the total stigmatic load of pollen was counted from one donor of each family (N=3).

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE, cache=FALSE,fig.height=4,fig.width=7, fig.cap="Barplot with the different treatments that provide information of the reproductive biology of the ten species. The y axis is the proportion of ovules converted to seed in percentage. The different, treatments which are presented in the legend are, hand cross pollination, hand self pollination, natural selfing and apomixis. More information about this treatments can be found in Methods and Appendices."}

load("r_biology.RData")

ggplot(bio,aes(x=Species,y=value,fill=variable))+
  geom_bar(stat="identity",position="dodge")+scale_fill_grey(start = 0, end = 0.85,name="Treatments") + 
  theme_bw()+xlab("Species")+ylab("Seed-setting rate (%)")+labs(title="Reproductive biology tests")+
  theme(axis.text.x = element_text(angle = 35, hjust = 1, face= "italic",family = "Arial Narrow"))+
  scale_y_continuous(minor_breaks = seq(0 , 100, 5), breaks = seq(0, 100, 10))
